Compiling Python Modules to Native Parallel Modules Using Pythran and OpenMP Annotations

### 2013-09-05 doc/papers/sc2013/paper.tex
Pierrick review

    HPC users traditionally rely on low-level, compiled language such as C or
    FORTRAN to perform computation-intensive tasks. As a consequence, it is a
    common situation to have HPC application written in a high-level language
    such as Python, calling native routines for computate-intensive tasks. To
    improve development speed and reduce maintenance costs, using a higher-level
    language like Python seems attractive. While it is usually associated with
    low performance, several solutions such as Cython, Numba, Parakeet or 
    Pythran offer to automatically or semi-automatically turn Python functions 
    into native ones.

@@

Parallel computations are supported by the Python standard library through the
\texttt{multiprocessing} module. It spawns several interpreters that
communicate through Inter Process Communication(IPC), using Python built-in
object serialization. This approach is only viable for computation intensive
independent tasks, since the communication and synchronization overheads are 
much greater than a light-weight threaded approach.

@@

While Python has a decorator mechanism\footnote{\see PEP 318,
\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
Python decorators.}, it only applies to functions, methods and classes and does
not make it possible to attach decorations to other statements.
PLW~\cite{dongara2007}, uses string instructions to hold such decorations. As
Python does not have anonymous block,~\footnote{\see PEP340,
    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
anonymous block support in Python.} one has to create a dummy \texttt{if 1:}
instruction to apply an annotation to a whole block. Pythran uses comments to
hold OpenMP directives, and internally stores them as string instructions.
Alternatively, the syntax \texttt{if 'my annotation':} is also supported.

Many OpenMP annotations are parametrized by clauses that list variables,
specifying their behavior with respect to parallel regions, e.g.\
\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
variables that have already been declared. However, there is no variable
declaration in Python, and all variables assigned in a function have the
function scope, called local scope and available through the \texttt{local()}
built-in. As a consequence, all variables that are referenced in a function are
considered when building such variable lists: there is no concept of variable
local to a block. Using the \texttt{default(none)} clause is possible to ensure
no variable gets forgotten when building such lists.

@@

For instance, in the motivating example, the \texttt{nsteps} variable is first
defined as argument, then read twice, which gives a def-use chain of
\{DEF, USE, USE\}. The associated nesting depths is
\{0, 1, 1\} so the variable scope is 0. \texttt{i} is defined as a loop iterator then
used in the loop body, which gives a chain of \{DEF, USE\}, with nesting
depths of \{2, 2\} and a scope of 2. Likewise \texttt{x} has a chain
of \{D, U\}, nesting depths of \{2, 2\} and a scope of 2, which means it can be
automatically declared local to the loop, thus not needing to be listed as
\texttt{private}.

@@

The extension of Python iterator to random access iterators is direct for the
standard containers: \texttt{list}\footnote{Python \texttt{list}s behave as C++
\texttt{vector}s.}, \texttt{set} or \texttt{dict}. Other iterators require more
care.

@@

The target machine has  8 AMD Opteron 6176 SE cores and 8 GO of RAM. It is
running a Linux kernel 3.10-2-amd64, x86\_64 GNU/Linux. The C++ compiler used is
\texttt{g++} version 4.8.1 and all applicationsa re compiled using the
\texttt{-Ofast} flag. Pythran generated programs are linked with Boost.Python
version 1.54. The Pythran version is extracted from the \texttt{sc2013} branch
of the git repository \url{https://github.com/serge-sans-paille/pythran}. Python
programs are run using the CPython implementation, version 2.7.5. Cython
programs are generated using version 0.19.1. Numba and Parakeet are installed
from the \texttt{master} branch of their respective repositories.

@@

@@

The best execution time out of five run is shown for each code version. If the
code does not compile with a particular compiler, the column is left blank. Two
code versions are shown for Pytrhan: one compiler without the OpenMP flag set
and one with the OpenMP flag set. The code itself is not modified, only the
compilation flag. A logarithmic scale is used due to the generally poor
performance of the pure Python version.

@@

Figure~\ref{fig:pb-pairwise} shows the execution time of the computation of a distance matrix between two random vectors, using a parallel nested loop. On that example, Pytrhan slightly outperforms Cython, parakeet and Numba. Taking into account the OpenMP annotation that flags the outermost loop as parallel yields an $\times 4.27$ speedup to Pythran.

>>>>

    High Performance Computing users traditionally rely on low-level,
    compiled language such as C or
    FORTRAN to perform computation-intensive tasks. As a consequence, it is a
    common situation to have High Performance Computing application written in a high-level language
    such as Python, calling native routines for computate-intensive tasks. To
    improve development speed and reduce maintenance costs, using a higher-level
    language like Python seems attractive. While it is usually associated with
    low performance, several solutions such as Cython, Numba, Parakeet or 
    Pythran offer to automatically or semi-automatically turn Python functions 
    into native ones.

@@

Parallel computations are supported by the Python standard library through the
\texttt{multiprocessing} module. It spawns several interpreters that
communicate through Inter Process Communication (IPC), using Python built-in
object serialization. This approach is only viable for computation intensive
independent tasks, since the communication and synchronization overheads are 
much greater than a light-weight threaded approach.

@@

While Python has a decorator mechanism\footnote{\see PEP 318,
\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
Python decorators.}, it only applies to functions, methods and classes and does
not make it possible to attach decorations to other statements.
PLW~\cite{dongara2007}, uses string instructions to hold such decorations. As
Python does not have anonymous block,\footnote{\see PEP340,
    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
anonymous block support in Python.} one has to create a dummy \texttt{if 1:}
instruction to apply an annotation to a whole block. Pythran uses comments to
hold OpenMP directives, and internally stores them as string instructions.
Alternatively, the syntax \texttt{if 'my annotation':} is also supported.

%what about flush clause which is not linked to a statement?
Many OpenMP annotations are parametrized by clauses that list variables,
specifying their behavior with respect to parallel regions, e.g.\
\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
variables that have already been declared. However, there is no variable
declaration in Python, and all variables assigned in a function have the
function scope, called local scope and available through the \texttt{local()}
built-in. As a consequence, all variables that are referenced in a function are
considered when building such variable lists: there is no concept of variable
local to a block. Using the \texttt{default(none)} clause is possible to ensure
no variable gets forgotten when building such lists.

@@

For instance, in the motivating example, the \texttt{nsteps} variable is first
defined as argument, then read twice, which gives a def-use chain of
\{DEF, USE, USE\}. The associated nesting depths is
\{0, 1, 1\} so the variable scope is 0. \texttt{i} is defined as a loop iterator then
used in the loop body, which gives a chain of \{DEF, USE\}, with nesting
depths of \{2, 2\} and a scope of 2. Likewise \texttt{x} has a chain
of \{DEF, USE\}, nesting depths of \{2, 2\} and a scope of 2, which means it can be
automatically declared local to the loop, thus not needing to be listed as
\texttt{private}.

@@

The extension of Python iterator to random access iterators is direct for the
standard containers: \texttt{list},\footnote{Python \texttt{list}s behave as C++
\texttt{vector}s.} \texttt{set} or \texttt{dict}. Other iterators require more
care.

@@

The target machine has  8 AMD Opteron 6176 SE cores and 8 GO of RAM. It is
running a Linux kernel 3.10-2-amd64, x86\_64 GNU/Linux. The C++ compiler used is
\texttt{g++} version 4.8.1 and all applications are compiled using the
\texttt{-Ofast} flag. Pythran generated programs are linked with Boost.Python
version 1.54. The Pythran version is extracted from the \texttt{sc2013} branch
of the git repository \url{https://github.com/serge-sans-paille/pythran}. Python
programs are run using the CPython implementation, version 2.7.5. Cython
programs are generated using version 0.19.1. Numba and Parakeet are installed
from the \texttt{master} branch of their respective repositories.

@@

%Ce n'est pas ce code qui ne donne pas les mêmes résultats avec Cython et Python et Pythran?

@@

The best execution time out of five run is shown for each code version. If the
code does not compile with a particular compiler, the column is left blank. Two
code versions are shown for Pythran: one compiler without the OpenMP flag set
and one with the OpenMP flag set. The code itself is not modified, only the
compilation flag. A logarithmic scale is used due to the generally poor
performance of the pure Python version.

@@

Figure~\ref{fig:pb-pairwise} shows the execution time of the computation of a distance matrix between two random vectors, using a parallel nested loop. On that example, Pythran slightly outperforms Cython, parakeet and Numba. Taking into account the OpenMP annotation that flags the outermost loop as parallel yields an $\times 4.27$ speedup to Pythran.
